```{r}
library(tidyverse)
setwd("~/Documents/Python/Datathon/MIT_COVID-19_Datathon")
```

nyt_data: county-level COVID statistics
```{r}
nyt_data = read_csv('~/Documents/Python/Datathon/covid-19-data/us-counties.csv') 
# https://github.com/nytimes/covid-19-data
```


Prison analysis
prison_df: state, prison name, and county
```{r}
prisons <- read_csv('prison_boundaries.csv')
prisons$VAL_DATE <- as.Date(prisons$VAL_DATE)
str_list <- paste(c('CORRECTION','PRISON'), collapse="|")
prison_df <- prisons %>% filter(grepl(str_list,NAME)) %>%
  select(NAME,STATE,COUNTY,POPULATION,CAPACITY,COUNTYFIPS) %>% 
  mutate(POPULATION = ifelse(POPULATION<0, NA, POPULATION)) %>%  # convert negative values to NA
  mutate(CAPACITY = ifelse(CAPACITY<0, NA, CAPACITY)) %>%  # convert negative values to NA
  mutate(fullness = POPULATION/CAPACITY) %>%
  group_by(COUNTYFIPS) %>% 
  mutate(prison_pop = sum(POPULATION)) %>%     # sum over all prisons in a county
  mutate(prison_capacity = sum(CAPACITY)) %>%  # sum over all prisons in a county
  mutate(prison_fullness = mean(fullness)) %>%  # calculate average prison fullness
  select(STATE,COUNTY,prison_pop,prison_capacity,prison_fullness,COUNTYFIPS) %>% unique()
  
prison_df
```

Get the states and their abbreviations, add State column to prison_df
```{r}
statenames <- read_csv('state_abbreviations.csv') %>% select(State,Code) %>% rename(STATE=Code)
prison_df <- merge(prison_df,statenames,all.x=TRUE)
prison_df
```

pr: state, fraction of prison population released
prison_counties_frac_released: counties with prisons, fraction of prisoners released statewide
frac_released = releases/prior_pop
```{r}
pr <- read_csv('Data/covid_prison_releases.csv') %>%
  rename(prior_pop = 'Population Prior to Releases',releases = "Overall Pop. Reduction / \nTotal Number of Releases") %>% 
  select(State,Facility,prior_pop,releases) %>% 
  filter(Facility=='Statewide') %>%              # not considering national-level releases
  mutate(releases = as.numeric(releases)) %>%   
  mutate(frac_released = releases/prior_pop) %>% # fraction released = # of releases / prior prison population
  select(State,frac_released) %>% drop_na()
pr

prison_counties_frac_released <- merge(prison_df, pr,by='State',all.x = TRUE) %>%
  mutate(frac_released = ifelse(is.na(frac_released),0,frac_released)) %>% # if a state has no reported releases, put 0 for frac_released
  select(COUNTYFIPS,frac_released,prison_pop,prison_capacity,prison_fullness,State) %>%unique()
prison_counties_frac_released
```


data for GLMs

```{r}
county_pop <- read_csv('new_county_demographics_df.csv') %>% rename(COUNTYFIPS = fips)
county_pop

nyt_data

nytdf <- nyt_data %>%
  mutate(prison_county = ifelse(fips %in% prison_counties_frac_released$COUNTYFIPS,1,0)) %>% # add column of 0/1: 1 if county has prison, 0 else
  rename(COUNTYFIPS = fips) %>%
  unique()
nytdf

nyt_with_prisons <- merge(nytdf, prison_counties_frac_released, all.x=TRUE) %>% group_by(COUNTYFIPS) %>%
  arrange(COUNTYFIPS,date) %>%
  mutate(diff_cases = cases-lag(cases,7)) %>%
  mutate(frac_change_cases = (cases-lag(cases,7))/lag(cases,7)) %>%
  filter(date >= as.Date('2020-03-01')) %>%
  select(COUNTYFIPS,county,state,date,cases,deaths,prison_county,frac_released,diff_cases,frac_change_cases,prison_pop,prison_capacity,prison_fullness) %>%
  unique()

prison_pop_df <- merge(nyt_with_prisons,county_pop,all.x=TRUE) %>% 
  group_by(COUNTYFIPS) %>%
  mutate(cases_per_capita = cases/tot_pop) %>%
  mutate(deaths_per_capita = deaths/tot_pop) %>%
  arrange(COUNTYFIPS,date) %>%
  mutate(cases_per_capita_7_days_ago = lag(cases_per_capita,7)) %>% filter(!is.na(diff_cases)) %>%
  mutate(prison_pop = ifelse(is.na(prison_pop),0,prison_pop))
prison_pop_df

prison_counties <- prison_pop_df %>% filter(prison_county==1)
prison_counties

```

```{r}
write.csv(prison_pop_df,'Data/full_population_and_prison_data.csv')
```


Models assessing the impact of having a prison in a given county
Model 1:
difference in cases (cases from today - cases from 7 days ago) ~ total county population, indicator function (contains a prison = 1, doesn't contain a prison = 0), county population density

Model 2:
difference in cases (cases from today - cases from 7 days ago) ~ total prison population (0 if no prison in county), total county population, cases per capita from 7 days ago, population density

```{r}
model1 <- lm(diff_cases ~ tot_pop + prison_county + population_density, data = prison_pop_df)
summary(model1)

model2 <- lm(diff_cases ~  prison_pop + tot_pop + cases_per_capita_7_days_ago + population_density, data = prison_pop_df)
summary(model2)
```

Models assessing the impact of release rates in counties with prisons


Model A:
Restricted to counties containing a prison
difference in cases (cases from today - cases from 7 days ago) ~ total county population, fraction of prisoners released statewide, county population density

Model B:
Restricted to counties containing a prison
difference in cases (cases from today - cases from 7 days ago) ~ total county population, fraction of prisoners released statewide, county population density, fraction of population in high risk age group

```{r}
prison_counties

modelA <- lm(diff_cases ~ tot_pop + frac_released + population_density, data = prison_counties)
summary(modelA)

modelB <- lm(diff_cases ~ tot_pop + frac_released +  population_density + lowrisk_agegroup_perc,data = prison_counties)
summary(modelB)

```

Repeat these analyses using frac_change_cases as the target variable instead of diff_cases

```{r}
model1 <- lm(frac_change_cases ~ tot_pop + prison_county + population_density, data = prison_pop_df)
summary(model1)

model2 <- lm(frac_change_cases ~  prison_pop + tot_pop + cases_per_capita_7_days_ago + population_density, data = prison_pop_df)
summary(model2)
```

```{r}
modelA <- lm(frac_change_cases ~ tot_pop + frac_released + population_density, data = prison_counties)
summary(modelA)

modelB <- glm(frac_change_cases ~ tot_pop + frac_released +  population_density + lowrisk_agegroup_perc,data = prison_counties)
summary(modelB)

```


```{r}
modelx <- lm(cases_per_capita ~ tot_pop + prison_county + population_density, data = prison_pop_df)
summary(modelx)

# this one has an R^2 value of 0.9!
modely <- lm(cases_per_capita ~  prison_pop + tot_pop + cases_per_capita_7_days_ago + population_density, data = prison_pop_df)
summary(modely)

modelz <- lm(deaths_per_capita ~ tot_pop + prison_county + population_density, data = prison_pop_df)
summary(modelz)

modelw <- lm(deaths_per_capita ~  prison_pop + tot_pop + cases_per_capita_7_days_ago + population_density, data = prison_pop_df)
summary(modelw)
```

Working with ACLU Massachusetts dataset: ACLU.csv gives for each considered DOC facility on each date the total population, detainees tested/positive, staff tested/positive, # released, and fips county code.

ACLU_cases dataframe gives date, county fips code, county-wide prisoner cases and cumulative sum of county-wide prisoner cases

mass_prison_counties adds information from prison_pop_df to ACLU_cases: we now have county-level cases, deaths, and population demographics

```{r}
# excluding hospitals, rehabilitation centers
DOC_Facility = c('MCI-Norfolk','NCCI-Gardn','OCCC','MCI-CJ','MCI-C','Pondville','MCI-F',
                       'MCI-Shirley','MCI-Shirley Min','MTC','SMCC','SBCC')
fips = c(25021,25027,25023,25021,25017,25021,25017,25017,25017,25023,25017,25017)
facilities_fips <- data.frame(DOC_Facility,fips)
facilities_fips

ACLU <- read_csv('ACLU_data.csv')
names(ACLU) <- str_replace_all(names(ACLU)," ","_")
ACLU <- merge(ACLU,facilities_fips,all.y=TRUE) %>% mutate(Date = as.Date(Date,"%m/%d/%y"))
ACLU <- ACLU[,c(1:11,27)] # remove empty columns
ACLU
write.csv(ACLU,'Data/ACLU_prison_data.csv')

ACLU_cases <- ACLU  %>% 
  select(Date,DOC_Facility,fips,`N_Positive_-_Detainees/Inmates`) %>%
  mutate(`N_Positive_-_Detainees/Inmates` = replace_na(`N_Positive_-_Detainees/Inmates`, 0))%>% # replace NA with 0 to get total counts
  group_by(fips,Date) %>%
  mutate(county_wide_prisoner_cases = sum(`N_Positive_-_Detainees/Inmates`)) %>% # sum over all cases in a county
  arrange(fips,Date) %>% 
  select(Date,fips,county_wide_prisoner_cases) %>%
  rename(COUNTYFIPS = fips,date = Date) %>% unique() %>% ungroup %>% group_by(COUNTYFIPS) %>%
  mutate(cumulative_county_wide_prisoner_cases = cumsum(county_wide_prisoner_cases)) # cumuylative cases in a county
ACLU_cases

mass_prison_counties <- merge(prison_pop_df,ACLU_cases,all.y=TRUE)
mass_prison_counties
```

```{r}
modelx <- lm(cases ~ cumulative_county_wide_prisoner_cases+ tot_pop, data = mass_prison_counties)
summary(modelx)
```


This chunk shows that across all prisons, we only have data on 10 releases --> not powerful enough for our model. We'll have to either A) consider case data that we get from each prison or B) consider jail data.

A) is likely not extremely useful given large gaps in testing
```{r}
tmp <- ACLU %>% select(fips,Date,N_Released) %>% mutate(N_Released = replace_na(N_Released,0))
sum(tmp$N_Released)
tmp <- read_csv('ACLU_data.csv')
names(tmp) <- str_replace_all(names(tmp)," ","_")
tmp <- tmp %>% select(Date,DOC_Facility,N_Released) %>% drop_na()
sum(tmp$N_Released)
```

Now considering ACLU data for county jails: MA_aclu_data has date, county info, population demographics, county-level cases and deaths, release statistics

```{r}
aclu_data <- read_csv('ACLU_data_Counties_All_DOC.csv') %>% rename(county = County,date = Date) %>%
  mutate(date = as.Date(date,"%m/%d/%y"))
aclu_data <- aclu_data[,1:19]
names(aclu_data) <- str_replace_all(names(aclu_data)," ","_")

aclu_data <- aclu_data %>% filter(county != 'DOC') %>% # just look at county jails, not DOC
  select(date,county,`N_Released_Pre-Trial`,N_Released_Sentenced)
aclu_data[is.na(aclu_data)] = 0

MA_prison_pop_df <- prison_pop_df %>% filter(state=='Massachusetts') %>%
  filter(county %in% aclu_data$county)

MA_aclu_data <- merge(MA_prison_pop_df,aclu_data)%>% group_by(county) %>% arrange(county,date)%>%
  mutate(cumulative_pretrial_release = cumsum(`N_Released_Pre-Trial`))%>%
  mutate(cumulative_sentenced_release = cumsum(N_Released_Sentenced)) %>%
  mutate(total_release = cumulative_sentenced_release + cumulative_pretrial_release) %>%
  select(date,county,COUNTYFIPS,tot_pop,cases,diff_cases,population_density,deaths,cumulative_sentenced_release,cumulative_pretrial_release,total_release,frac_change_cases)

prison_pop_df
MA_aclu_data
```
```{r}
modelx <- lm(frac_change_cases ~ total_release + tot_pop, data = MA_aclu_data)
summary(modelx)
```

Notes: I think the way I calculate R^2 is wrong